Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-29 survey/survey.tex
minor fix to sec 2
starting data ii

Bijections on \fin{n} have a specific name: permutations. As is well-known
permutations can be generated by sequences of transpositions. Thus we
can create a small language for writing permutations as
\[\begin{array}{rcl}
p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p \fatsemi p
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

Although this language appears weak, it is universal for boolean
combinational circuits. The result follows from the following
construction of the Toffoli gate which is universal for combinational
circuits. We write $[2^n]$ for the finite set representing $n$-bit
words with the canonical ordering for binary numbers. Thus when $n=3$,
the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to
the 3-bit words $\{000,001,010,011,100,101,110,111\}$. The Toffoli
gate is a reversible function $[2^3] \rightarrow [2^3]$ which swaps
$110$ and $111$, i.e., it can be realized as $\mathit{swap}~6~7$.

In programming practice we are however more interested in structured
data which will be the subject on next section. What we learn from
this short investigation using untyped flat sets is what the
\emph{purely operational} view of the theory would be. In particular,
it tells us that permutations are an inescapable part of the fabric of
reversible computing.  However as permutations are untyped, and act on
the canonicalized version of $n$-element sets (i.e. those sets where
all the structure has been forgotten), these are a rather pale shadow
of the rich tapestry of information-perserving transformation of
structured data which we investigate next.

@@

The elementary building blocks of type theory are the empty type
($\bot$), the unit type ($\top$), and the sum ($\uplus$) and product
($\times$) types.  These have a nice correspondance (Curry-Howard)
to logic:

>>>>

Bijections on \fin{n} have a specific name: permutations. As is
well-known permutations can be generated by sequential compositions of
transpositions. Thus we can create a small language for writing
permutations as
\[\begin{array}{rcl}
p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p \fatsemi p
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

Although this language appears weak, it is universal for boolean
combinational circuits. The result follows from the following
construction of the Toffoli gate and the fact that any combinational
circuit can be realized using compositions of Toffoli gates. We write
$[2^n]$ for the finite set representing $n$-bit words with the
canonical ordering for binary numbers. Thus when $n=3$, the finite set
has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to the 3-bit words
$\{000,001,010,011,100,101,110,111\}$. The Toffoli gate is a
reversible function $[2^3] \rightarrow [2^3]$ which swaps $110$ and
$111$, i.e., it can be realized as $\mathit{swap}~6~7$.

In programming practice we are however more interested in structured
data which will be the subject on next section. What we learn from
this short investigation using untyped flat sets is what the
\emph{purely operational} view of the theory would be. In particular,
it tells us that permutations are an inescapable part of the fabric of
reversible computing.  However as permutations are untyped, and act on
the canonicalized version of $n$-element sets (i.e. those sets where
all the structure has been forgotten), these are a rather pale shadow
of the rich tapestry of information-perserving transformations of
structured data which we investigate next.

@@

Instead of the spaces consisting solely of unstructured isolated
points, we now investigate structured spaces built from sums and
products of elementary spaces. This structure corresponds to the
building blocks of type theory which are: the empty type
($\mathbf{0}$), the unit type ($\mathbf{1}$), the sum type ($\uplus$),
and the product ($\times$) type. Before getting into the formal
theory, let's consider the space
$(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
\mathbf{1})$. The subspace $(\mathbf{1} \uplus \mathbf{0})$ is the sum
of the space $\mathbf{1}$ containing one element and the empty space
$\mathbf{0}$. By construction this space is different from just the
space $\mathbf{1}$ but one can argue that the spaces are equivalent in
some sense.

\amr{new wavefront for new pass}

These have a nice correspondance (Curry-Howard) to logic:
